---
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography: rs15_unclassified.bib
link-citations: true
params:
  j: 15
---


```{r include = FALSE}
library(tidyverse)
library(leaflet)

load('assets/cases_db.Rda')
cases <- dat
dat <- read_csv('assets/generic_types_RSDB_new.csv') |> 
    janitor::clean_names()

x <- dat$id[params$j]
rs <- dat |> filter(id== x)
cs <- cases |> filter(id== x)
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```

# Salt marshes to tidal flats

**Main contributors**: Steven Alexander, Sean Fagan

**Other contributors**: Reinette (Oonsie) Biggs

Last update: 2026-06-26

```{r map, out.width='100%', out.height='300px', results = 'markup'}


cs <- cases |> 
    filter(type == rs$regime_shift_name) 

m <- leaflet(cs) |> 
    addTiles('https://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png') |> 
    addCircleMarkers(
        lng = ~long, lat=~lat, radius = 2, 
        color = 'orange', 
        popup = ~popups) |> 
    setView(5,10, zoom = 2) 
m
```

### Summary

Estuaries and salt marshes provide more ecosystem services per unit area than any other ecosystem worldwide. However, these valuble ecosystems are under pressure from sea level rise and invasive species. The loss of ecosystem services from this regime shift includes: pollution filtration, storm protection, fisheries enhancement, and carbon sequestration. Salt marshes have long been recognized for their ability to keep pace with rising sea levels, regulating their own elevation through a series of non-linear biophysical feedback mechanisms. But this self-regulation has limits. Beyond certain thresholds in the rates of sea level rise (RSLR) and sediment delivery, these feedback mechanisms can no longer keep pace [@kirwan2010]. There are wide-ranging management strategies, such as invasive species removal and sediment supply restoration. Across all strategies, the aim is to retain or increase native vegetation establishment [@zhao2021].

##  {.tabset}

### Categorical attributes

::::: {style="display: flex;"}
::: {style="width: 50%;"}
#### Impacts

**Ecosystem type:** Marine & coastal

**Key ecosystem processes:** Soil formation, Primary production, Nutrient cycling

**Biodiversity:** Biodiversity

**Provisioning services:** Fisheries, Fuel and fiber crops

**Regulating services:** Climate regulation, Water purification, Regulation of soil erosion, Natural hazard regulation

**Cultural services:** Recreation, Aesthetic values, Knowledge and educational values

**Human well-being:** Livelihoods and economic activity, Cultural, Aesthetic and recreational values

**Links to other regime shifts:** NA
:::

::: {style="width: 50%;"}
#### Drivers

**Key drivers:** Vegetation conversion and habitat fragmentation, Infrastructure development (e.g. roads, pipelines), Species introduction or removal, Soil erosion & land degradation, Global climate change

**Land use:** NA

#### Key attributes

**Spatial scale:** Local/landscape (e.g. lake catchment, community)

**Timescale:** Years, Decades

**Reversibility:** Hysteretic (difficult to reverse)

**Evidence:** Models, Paleo-observation, Contemporary observations, Experiments

**Confidence: existence of the regime shift:** Contested – Reasonable evidence both for and against the existence of RS

**Confidence: mechanisms underlying the regime shift:** Well established – Wide agreement on the underlying mechanism
:::
:::::

### Detail info and key references

#### Alternative Regimes

Coastal ecosystems within the intertidal zones of temperate regions can undergo regime shifts between a vegetated salt marsh state and an unvegetated flat. These regimes are alternative stable states that emerge from the coupled interaction of biological and physical processes [@marani2010]. Which alternate regime emerges varies regionally and depends on a range of interacting factors, including tidal basin topography, climate, hydrology, vegetation cover, and sediment loading. The two alternate regimes are:

##### Salt Marsh

This regime is dominated by dense, largely monospecific stands of Spartina alterniflora (smooth cordgrass) along the east coast of North America and Spartina anglica (common cordgrass) in continental northern Europe, halophytic grasses adapted to saline, waterlogged conditions and frequent tidal inundation [@marani2010]. These high-density swards occupy the intertidal zone, where they are flooded daily. Spartina is generally found between mean sea level and mean high tide, but its precise elevational range and the optimum elevation of the marsh platform vary regionally with tidal range, nutrient loading, sediment supply, and climate [@mckee1988; @morris2002]. Beyond their structural dominance, salt marshes mediate the exchange of water, sediment, and nutrient fluxes in coastal landscapes, sequester organic carbon at high rates, and host diverse coastal biota [@finotello2020].

##### Tidal Flat / Subtidal Flat

These regimes are devoid of vascular vegetation. Instead, they are characterized by microbially driven food webs dominated by benthic macroinvertebrates and benthic microalgae [@alongi1998]. Tidal flats sit just below mean sea level (MSL) and are exposed only during the lowest tides [@fagherazzi2006; @defina2007], whereas subtidal flats lie permanently submerged below the low tide line. The conceptual basis for the regime shift rests on this elevation distinction: in microtidal systems, marsh platforms sit above MSL while flats sit below it, so the loss of marsh volume above MSL is what ultimately triggers the transition [@ganju2017].

#### Drivers and Causes

##### Shift from Salt Marsh to Tidal Flat / Subtidal Flat

The primary cause of this regime shift is sea level rise. The link between climate change and accelerating sea level rise, along with its consequences for coastal marine systems, is well established [@unep2006]. Salt marshes have long been able to regulate their own elevation in response to rising seas through a suite of non-linear biophysical feedbacks [@murray2008; @kirwan2010]. However, a threshold exists in the rate of sea level rise (RSLR) beyond which these self-regulating mechanisms can no longer keep pace [@kirwan2010]. For example, if reduced sediment supply from upstream inhibits soil accretion on the marsh platform, this undermines its ability to stay above rising sea levels. While a marsh receiving abundant sediment can persist under relatively rapid sea level rise. This is reinforced by a positive feedback loop: as marsh area is lost, the remaining vegetation traps less sediment, which accelerates further degradation [@donatelli2018].
The second pathway is a shift toward consumer control, in which unchecked populations of herbivores such as snails and crabs come to govern ecosystem productivity and structure. This top-down dominance can arise from the overharvesting of predators or the introduction of invasive or exotic species. Additional stressors, including nitrogen loading and the weakening of plant defenses under climatic extremes such as elevated salt stress, can also promote consumer control, triggering the cascading vegetation losses that push the system toward a tidal or subtidal flat regime [@bertness2008].
A third pathway is lateral erosion where stronger waves cause faster edge retreat [@finotello2020]. If the adjacent tidal flat widens past a critical point, the resulting increase in wave power erodes the bed further, widening the flat even more. This feedback loop can lead to irreversible marsh collapse [@fagherazzi2015]. A related but distinct pathway is interior collapse, where internal fragmentation and ponding cause the marsh to lose elevation from within, rather than from the edges [@schepers2020].

#### Recovery and Reverse Transitions

##### Shift from Tidal Flat / Subtidal Flat back to Salt Marsh

While much of the literature emphasizes the marsh-to-flat transition, regime shifts in these systems are not always unidirectional, and under certain conditions tidal flats can revert to vegetated marsh. This reverse transition has been documented in the Greater Thames estuary in England, where van der Wal and Pye [-@vanderwal2004] describe a salt marsh cycle in which lateral retreat at the marsh edge is accompanied by a gradual rise in mudflat elevation. Once the flat reaches a sufficient elevation, Spartina anglica recolonizes, lateral erosion slows, and the marsh effectively rejuvenates [@bacino2020]. Such recovery is, however, contingent on a narrow set of biological and physical conditions. Effective seed dispersal from neighboring marshes is critical since recolonization cannot proceed without an adequate supply of viable propagules reaching the flat surface [@zhao2021]. Sufficient sediment supply is equally essential, both to raise the flat into the elevation range that Spartina can tolerate and to sustain accretion once vegetation is established [@fagherazzi2015]. Where wave energy at the marsh edge is high, recovery may also depend on the natural or restored presence of wave-attenuating features such as oyster reefs, which reduce the hydrodynamic stress that would otherwise inhibit seedling establishment [@mcloughlin2015]. It is important to emphasize that not all forms of marsh degradation are reversible on management-relevant timescales.

#### Impacts on Ecosystem Services and Human Well-being

##### Shift from Salt Marsh to Tidal Flat / Subtidal Flat

Estuaries and salt marshes provide more ecosystem services per unit area than any other ecosystem worldwide [@unep2006]. Since prehistoric times, marshes have supplied edible plants and animals, thatch and fiber for building, fodder for livestock, and fuel for cooking. More recently, they have been recognized for storm protection [@costanza2008], biogeochemical filtering [@drake2009], carbon sequestration [@chmura2003], and as nurseries for commercially harvested finfish and shellfish [@gedan2009]. Tidal and subtidal flats also provide ecosystem services, often performing similar functions through different ecological pathways. Tidal flats, for example, support mollusk and crustacean populations that sustain their own distinct fisheries [@unep2006]. Despite this functional overlap, the shift from a vegetated marsh to an unvegetated flat results in a substantial net loss of key services, including pollution filtration, storm protection, and fisheries enhancement [@gedan2009]. The climate consequences are also considerable: ongoing global marsh loss represents a meaningful release of previously sequestered carbon [@campbell2022].
The erosion of these services has direct consequences for human well-being. Coastal livelihoods are diminished as fisheries productivity declines, public health is compromised by the reduced capacity of the system to filter pollutants and excess nutrients, and vulnerability to climate change increases as communities lose both natural storm buffering and the broader adaptive capacity that healthy marshes provide.

#### Management options

Addressing the rate of sea level rise requires sustained international commitment to climate change mitigation, and because the global climate system responds to policy interventions only over decadal to centennial timescales, any strategy targeting RSLR must adopt a long-term outlook. At more local and regional scales, however, a range of management actions can be implemented to address the more tractable drivers of regime shift: sediment delivery, lateral erosion, and consumer control. Mudd [-@mudd2011] suggests that in watersheds where rivers have been impounded, scheduled dam releases could be coordinated to deliver pulses of sediment that promote marsh establishment and expansion. Where lateral erosion is the dominant pathway, nature-based interventions such as the conservation or construction of oyster reefs adjacent to marsh edges can dissipate wave energy and reduce edge retreat [@mcloughlin2015]. On the consumer control side, restoration practices that remove invasive or exotic species and reestablish native predator populations can shift community structure in ways that suppress runaway grazing and allow vegetated marsh to recover [@gedan2009; @zhao2021].
Effective management must also account for the broader context in which salt marshes exist. These systems occupy a heterogeneous coastal zone that is rapidly becoming human-dominated, with roughly one-third of the global human population now living in coastal areas [@unep2006]. This proximity imposes constraints that shape what restoration can realistically achieve. The long-term capacity of a marsh to maintain its elevation under accelerating sea level rise depends on the unimpeded flow of water, sediment, and nutrients into and out of the system, yet tidal restrictions caused by roads, railway crossings, and undersized culverts routinely disrupt these exchanges and have been linked to shifts in community composition [@gedan2009]. A second key constraint is the encroachment of development on the marsh's upland boundary, which leaves no accommodation space for the platform to migrate landward as sea levels rise. Where this migration is permitted, however, upland drowning can offset some of the marsh loss caused by RSLR, as documented in Chesapeake Bay, where massive upland-to-wetland conversion has historically compensated for marsh erosion [@schieder2018]. Where the full suite of pressures operating across spatial and temporal scales is not addressed in an integrated way, the long-term resilience of the system can be compromised, and restoration efforts undermined.

#### Feedback mechanisms

##### Vertical accretion feedback / Marsh elevation–sedimentation feedback (local, well-established, negative feedback)

This feedback mechanism is often considered central to the long-term stability of salt marshes under sea level rise. Marsh elevation relative to mean sea level controls inundation frequency and depth, which in turn controls the volume of waterborne sediment delivered to the marsh platform during each tidal flood. As relative sea level rises, the marsh platform is flooded more frequently and more deeply, increasing the sediment available for deposition on the surface. This elevated deposition raises the platform's elevation, which then reduces the relative degree of submergence again, restoring the system toward equilibrium with sea level [@kirwan2010; @morris2002].
This is a negative, self-correcting feedback. A marsh experiencing a small elevation deficit accretes faster, while a marsh that is already keeping pace accretes more slowly, so the loop dampens deviations from equilibrium rather than amplifying them. This mechanism also has a limit: it can only restore equilibrium up to a threshold rate of relative sea level rise, beyond which sediment delivery can no longer match the pace of submergence [@kirwan2010]. Beyond this threshold, the negative feedback effectively saturates, and the marsh platform begins to accumulate a persistent elevation deficit.

##### Vegetation–sediment trapping feedback (local, well-established, positive feedback)

Closely linked to the accretion feedback above, this mechanism concerns the relationship between vegetation cover and the marsh's capacity to capture incoming sediment. Dense vegetation cover traps sediment effectively, and this deposition supports continued plant growth by maintaining suitable elevation, which in turn sustains the dense vegetation cover needed to trap further sediment. Under degraded conditions, however, this same relationship reverses into a destabilizing loop. When marsh area and vegetation are lost, the remaining vegetation traps less sediment, which reduces accretion and accelerates further degradation. This is a positive, self-reinforcing feedback: each cycle of vegetation loss makes the next cycle of loss more likely, which is why marsh dieback tends to accelerate once it begins rather than stabilizing at a partial loss.

##### Lateral erosion–tidal flat widening feedback (local, well-established, positive feedback)

Stronger waves cause faster edge retreat at the marsh boundary [@finotello2020]. If the adjacent tidal flat widens past a critical point, the resulting increase in wave power erodes the bed further, widening the flat even more [@fagherazzi2015]. This is a strongly self-reinforcing positive feedback. Once the threshold is exceeded, the loop becomes self-sustaining and can lead to irreversible marsh collapse, even in the absence of further sea level rise [@fagherazzi2015].

##### Consumer control / herbivory feedback (local, well-established, positive feedback)

Unchecked populations of herbivores such as snails and crabs can come to govern ecosystem productivity and structure, a shift that can arise from the overharvesting of predators or the introduction of invasive or exotic species. Additional stressors, including nitrogen loading and the weakening of plant defenses under climatic extremes such as elevated salt stress, can also promote this shift toward consumer control [@bertness2008]. Once consumer control is established, grazing pressure drives cascading vegetation loss, which removes the structure that would otherwise help limit herbivore access and impact. This loss of vegetative structure leaves the platform more exposed to continued grazing, reinforcing the cycle and pushing the system toward a tidal or subtidal flat regime [@bertness2008]. This operates as a positive feedback in its destabilizing direction, since each round of grazing-driven loss compounds the conditions for further loss.

##### Reverse-transition accretion feedback (local, well-established, negative feedback)

This feedback operates on the recovery pathway from tidal flat back to vegetated marsh, and mirrors the vertical accretion feedback described above, but acting on bare sediment rather than an existing platform. Van der Wal and Pye [-@vanderwal2004] describe a salt marsh cycle in which lateral retreat at the marsh edge is accompanied by a gradual rise in mudflat elevation. Once the flat reaches a sufficient elevation, Spartina anglica recolonizes, and the marsh effectively rejuvenates [@bacino2020]. Established vegetation then begins trapping additional sediment, accelerating further accretion and slowing the lateral erosion that drove the original collapse.

#### Drivers

Sea level rise (regional, well established): Rising seas reduce the elevation of capital marsh platforms relative to mean sea level, and once the rate of sea level rise outpaces the marsh's capacity for vertical accretion, the system crosses a threshold beyond which self-regulating feedbacks can no longer maintain marsh elevation [@kirwan2010].
Reduced sediment supply (regional, well-established): When upstream dams, watershed alterations, or reduced fluvial inputs limit the sediment reaching the marsh platform, soil accretion cannot keep pace with rising seas, weakening the marsh's ability to maintain elevation above mean sea level [@fagherazzi2015].
Vegetation loss and habitat fragmentation (local, well-established): The loss of marsh vegetation reduces sediment trapping and organic matter accumulation on the platform, reinforcing a positive feedback loop in which further vegetation die-off accelerates erosion and elevation loss [@donatelli2018].
Lateral erosion (local, well-established): Wave action at the marsh edge drives progressive retreat of the marsh boundary, and as the adjacent flat widens, increased wave power further erodes the bed, creating a self-reinforcing cycle that can lead to irreversible marsh collapse [@fagherazzi2015].
Interior collapse (local, well-established): Internal fragmentation and ponding within the marsh interior cause localized waterlogging and vegetation dieback, leading to elevation loss from within the marsh rather than from its edges [@schepers2020].
Consumer control/herbivory (local, well-established): Unchecked populations of herbivores such as snails and crabs, often resulting from predator overharvesting, can drive extensive vegetation dieback that destabilizes the marsh platform and accelerates its transition toward an unvegetated flat [@bertness2008].
Invasive species (regional, well-established): The introduction of non-native species can disrupt existing predator-prey dynamics and competitive relationships, often promoting consumer control or altering sediment and vegetation dynamics in ways that destabilize the marsh [@bertness2008].

#### Key thresholds

Rate of relative sea level rise (RSLR) (global, well-established): Marshes can generally keep pace with sea level rise up to roughly 7–12 mm/year, depending on local sediment supply and estuary type, beyond which accretion can no longer offset submergence and elevation deficits accumulate [@wu2017].
Estuary-specific RSLR threshold (regional, well-established): The critical rate of sea level rise varies by hydrological setting, with riverine-dominated estuaries tolerating higher rates (~10.3 mm/yr) than marine-dominated estuaries (~7.2 mm/yr) due to differences in mineral sediment supply from rivers [@wu2020].
Rate-induced critical transition (local, proposed): Marsh collapse can be triggered by how quickly the sea-level rise rate changes, since a sufficiently fast acceleration can push the system past a tipping point even before the long-term equilibrium threshold is reached [@fagherazzi2020].
Critical tidal flat width (local, proposed): Once an adjacent tidal flat exceeds a threshold width, irreversible marsh erosion occurs even without any sea level rise, driven by self-reinforcing feedbacks among flat widening, flat deepening, and local wind-wave generation [@fagherazzi2015].
Sediment concentration / turbidity threshold (regional, proposed): Maintenance of tidal flats requires a minimum suspended sediment turbidity that increases with tidal range; falling below this threshold destabilizes the flat and, by extension, the adjacent marsh edge [@grandjean2024].
Wave-energy threshold at the marsh edge (local, contested): While flat-width thresholds are well supported, evidence on whether a critical wave-power threshold exists at the marsh edge itself is mixed, with some global analyses finding a linear (non-threshold) relationship between wave energy and erosion rate [@leonardi2016; @fagherazzi2020].

#### References

