---
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography: rs3_unclassified.bib
link-citations: true
params:
  j: 3
---


```{r include = FALSE}
library(tidyverse)
library(leaflet)

load('assets/cases_db.Rda')
cases <- dat
dat <- read_csv('assets/generic_types_RSDB_new.csv') |> 
    janitor::clean_names()

x <- dat$id[params$j]
rs <- dat |> filter(id== x)
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```

# Coral Reef Regime Shifts

**Main contributors**: Juan Carlos Rocha

**Other contributors**: Reinette (Oonsie) Biggs, Garry Peterson, Albert Norström, Sean Fagan

Last update: 2026-04-14
```{r map, out.width='100%', out.height='300px', results = 'markup'}


cs <- cases |> 
    filter(type == rs$regime_shift_name) 

m <- leaflet(cs) |> 
    addTiles('https://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png') |> 
    addCircleMarkers(
        lng = ~long, lat=~lat, radius = 2, 
        color = 'orange', 
        popup = ~popups) |> 
    setView(5,10, zoom = 2) 
m
```


### Summary

<p>Regime shifts in coral reefs typically involve a change in species dominance from hard corals (3D structure) to algal dominance. Less commonly documented shifts include shifts from hard corals to soft coral dominance, corallimorpharians, urchin barrens, or sponge dominance. All these regime shifts result in loss of diversity and structural complexity, and are typically triggered by a combination of overfishing, pollution, diseases, and climate change. Loss of biodiversity and coral bleaching make coral systems more vulnerable to such stressors.</p>
##  {.tabset}

**Type of regime shift**: Coral-dominated reef to algal-dominated reef / soft coral, corallimorpharian, sponge, or urchin barren state

**Ecosystem type**:

- Marine & coastal

**Land uses**:

- Fisheries
- Conservation
- Tourism

**Spatial scale**:
- Local/landscape (e.g. lake, catchment, community)

**Continent or Ocean**:
- Multiple tropical oceans

**Region**: Tropical and subtropical coral reef systems

**Countries**: Multiple

{.tabset}
--------------------------------------------------

### Categorical attributes

::::{style='display: flex;'}

:::{style='width: 50%;'}

#### Impacts

**Ecosystem type**:

- Marine & coastal

**Key ecosystem processes**:

- Soil formation

**Biodiversity**:

- Biodiversity

**Provisioning services**:

- Fisheries
- Wild animal and plant products

**Regulating services**:

- Water purification
- Regulation of soil erosion
- Pest & disease regulation
- Natural hazard regulation

**Cultural services**:

- Recreation
- Aesthetic values
- Knowledge and educational values
- Spiritual and religious

**Human well-being**:

- Food and nutrition
- Livelihoods and economic activity
- Security of housing & infrastructure
- Cultural
- Aesthetic and recreational values

**Links to other regime shifts**:

- Freshwater eutrophication
- Forest to savanna
- Fisheries collapse
:::

:::{style='width: 50%;'}

#### Drivers

**Key drivers**:

- Harvest and resource consumption
- External inputs (e.g. fertilizers, pest control, irrigation)
- Adoption of new technology (e.g. new fishing nets)
- Species introduction or removal
- Disease
- Soil erosion & land degradation
- Environmental shocks (e.g. fire, floods, droughts)
- Global climate change

**Land use**:

- Fisheries
- Conservation
- Tourism


#### Key attributes

**Spatial scale**:

- Local/landscape (e.g. lake catchment, community)

**Time scale**:

- Years
- Decades

**Reversibility**:

- Hysteretic (difficult to reverse)

**Evidence**:

- Models
- Paleo-observation
- Contemporary observations
- Experiments

**Confidence: existence of the regime shift**:

- Well established – Wide agreement in the literature that the RS exists

**Confidence: mechanisms underlying the regime shift**:

- Well established – Wide agreement on the underlying mechanism

:::

::::

### Detail info and key references

#### Alternative regimes

Regime shifts in coral systems are usually associated with changes in the species dominance of this ecological community and consequent changes in ecosystem structure. Coral reefs are marine ecosystems, three-dimensional shallow-water structures dominated by scleractinian or hard corals [@bellwood2004]. The most well-documented regime shifts entail shifts from hard coral to fleshy seaweed (macroalgae) dominance. However, shifts from hard coral to corallimorpharian, soft coral, sponge, and urchin barren states have also been documented [@norstrom2009], along with less frequently reported shifts toward algal turfs, benthic cyanobacterial mats, and zoanthids [@crisp2022].

The frequency with which different alternative regimes appear in the literature may not reflect their true frequency in nature. Coral-to-macroalgae transitions dominate the published record, but this likely reflects research design as much as ecology: most studies have surveyed only two or three benthic groups (typically coral and macroalgae), concentrated in the Caribbean, and examined reefs that were initially coral-dominated, which structurally limits the detection of other kinds of transitions [@crisp2022]. Some regime shifts may also be “cryptic” — that is, missed by conventional monitoring that focuses on hard corals and macroalgae while overlooking other benthic taxa such as octocorals, which can increase in abundance even as scleractinian cover declines [@tsounis2017].

##### Coral-dominated reefs

Coral reefs are marine communities considered one of the most biodiverse and economically important ecosystems on the planet [@hoeghguldberg2007]. Coral reefs are old calcareous three-dimensional structures created by corals — coral polyp colonies in symbiosis with microscopic algae. This regime implies high biodiversity and three-dimensional habitat complexity that offers shelter to other species.

Coral reefs occupy less than 0.5% of the ocean floor but sustain up to 25% of marine biodiversity, and produce at least 10% of fish consumed by humans [@moberg1999]. Due to their three-dimensional structure, it is estimated that there are over 800,000 plant and animal species on coral reefs globally [@moberg1999; @fisher2015]; this rich diversity makes coral reefs a target for tourism. In over 100 countries, coral reefs are major natural and economic resources [@goreau1994]. For example, coral reefs are one of the major attractions in the Caribbean. In that region, tourism is a major source of foreign currency, and in some Caribbean countries tourism accounts for up to half of the nation’s gross domestic product [@hoeghguldberg2007].

Coral reefs can also reduce coastal erosion by reducing the impact of ocean currents and waves. Consequently, intact coral reefs protect coastal infrastructure and valuable sandy beaches. Furthermore, increases in erosion can result in the degradation or loss of other coastal ecosystems such as mangroves and seagrass areas, which provide valuable ecosystem services such as fish nurseries [@moberg1999; @hoeghguldberg2007].

##### Algae-dominated reefs

Algae-dominated reefs (more specifically, macroalgae-dominated reefs) emerge when fleshy macroalgae overgrow corals and come to dominate the benthos. While turf algae and crustose coralline algae are natural and often beneficial components of healthy reefs, it is the proliferation of fleshy macroalgae that drives the loss of habitat complexity characteristic of degraded systems [@bellwood2004]. Macroalgae-dominated reefs have lower biodiversity and fish production, reduced primary production, and provide less protection from coastal erosion [@moberg1999].

**Macroalgae-dominated reefs are maintained by ecological feedbacks:** macroalgae can prevent coral settlement, reduce coral fecundity, and, in some cases, cause microbially mediated coral mortality, making recovery difficult without major changes in the underlying drivers of decline [@graham2015]. However, such regimes are not always permanent. Natural reversals back to coral dominance have been documented, sometimes triggered by severe heat stress, causing macroalgal die-off in systems where herbivory remains high [@graham2024]. This challenges the view that coral-to-macroalgae regime shifts are necessarily locked in by hysteresis.

##### Corallimorpharians, soft corals, and sponges

Many regimes can dominate former coral reefs when they are dramatically disturbed [@bruno2009; @norstrom2009]. Norström et al. [-@norstrom2009] report that systems dominated by corallimorpharians, soft corals (octocorals), and sponges are alternative regimes that coral reefs can shift to when strongly disturbed. These taxa are thought to reduce habitat complexity and hence biodiversity, and fisheries production and tourism services may be reduced as a result.

Shifts toward octocoral-dominated communities have been increasingly observed, particularly in the Caribbean, where gorgonian abundance has risen on reefs with declining scleractinian cover. Octocorals appear more resilient than hard corals to disturbances such as hurricanes, bleaching, and disease, and can rapidly colonise space left by dead corals [@tsounis2017]. Corallimorph-dominated shifts have also been documented, often linked to localised disturbances such as shipwrecks, where iron leaching from metal structures is thought to stimulate corallimorpharian growth [@work2008].

##### Urchin barrens

Urchins are herbivores that live on coral reefs. In the absence of predatory fish, urchins can become the dominant grazer on reefs and produce urchin barrens. Urchin barrens inhibit coral growth because urchins erode calcium carbonate from living and dead corals, eroding reef structure. When urchins are present at densities high enough that their erosion exceeds coral growth, urchin barrens can form [@norstrom2009].

Urchin barrens are the regime with the most severe habitat complexity loss, which reduces habitat for other species. There is low production of fish, and the urchin populations are vulnerable to disease. However, urchins can be a valuable species when harvested, and so urchin barrens can support an urchin fishery.

#### Drivers and causes of the regime shift

Coral reef regime shifts typically emerge from the interaction of multiple stressors operating across spatial and temporal scales, rather than from single, isolated disturbances [@bellwood2004; @hughes2017]. These shifts occur when pressures exceed the capacity of reef systems to absorb disturbance and maintain their structure and function. The main drivers can be grouped into acute shocks (such as marine heatwaves and the coral bleaching they cause, tropical cyclones, and disease outbreaks) and chronic pressures, including overfishing (particularly of herbivores), nutrient and sediment runoff from land use, coastal development, and the long-term warming and acidification associated with climate change. Systematic reviews indicate that coral bleaching, tropical cyclones, and reduced herbivory are the most frequently cited proximate drivers, together accounting for roughly half of reported cases [@crisp2022].

Hard coral–dominated reefs are generally considered more desirable than alternative states because of their important biological, social, and economic roles. When reefs shift away from hard coral dominance, the same broad set of drivers (bleaching, storms, overfishing, and land-based pollution) is usually involved, but the type of new regime that develops depends on local conditions and the sequence of events. Some Indian Ocean and broader Indo-Pacific reefs have shifted, or are vulnerable to shifting, toward alternative benthic states such as corallimorpharian, soft-coral, sponge, or algal dominance rather than following a single coral-to-macroalgae pathway [@norstrom2009]. Herbivory is one important mechanism shaping these outcomes, because grazing by herbivorous fishes and urchins can suppress macroalgae and maintain space for coral recruitment [@adam2015; @edwards2014]. However, the final trajectory depends on local combinations of bleaching, fishing, nutrient loading, storm damage, structural complexity, and recovery timing.

Whether a reef recovers after a disturbance or shifts to an alternative state also depends on its initial condition. Factors such as structural complexity, water depth, the number of juvenile corals, herbivorous fish biomass, and nutrient levels all influence resilience. In contrast, simply being designated as a no-take marine reserve does not, on its own, strongly determine the reef’s recovery path [@graham2015].

##### Shocks contributing to the regime shift

**Thermal anomalies (regional, well-established):** Thermal anomalies produce coral bleaching, a phenomenon seen when corals release their zooxanthellae due to thermal stress, leading to the loss of coral color. If temperatures exceed summer maxima by 1–2°C for over 3 weeks, then bleaching will occur, with more severe bleaching as thermal anomalies intensify and lengthen [@hoeghguldberg2007]. Berkelmans et al. [-@berkelmans2004] found that in the Great Barrier Reef, Australia, the maximum sea surface temperature (SST) over any 3-day period during the bleaching season (summer) predicted the presence/absence of bleaching with 73.2% accuracy. Massive events of coral bleaching have been recorded more recently [@berkelmans2004]. According to model predictions, coral bleaching events are expected to become increasingly frequent and severe in the coming decades due to global warming [@wooldridge2005; @hoeghguldberg2007]. Corals may survive and recover from bleaching after mild thermal stress, but typically show reduced growth, calcification, and fecundity and may experience greater incidences of coral disease [@bruno2007; @hoeghguldberg2007].

**Hurricanes (regional, contested):** Hurricanes are shock events that can destroy coral structure, reducing coral populations. These shocks can provide opportunities for algal populations to dominate over coral reefs, especially when these shocks occur in synergy with other drivers. However, hurricanes might also have a beneficial role when passing several hundred kilometers from coral reefs as they potentially cool down water from thermal anomalies [@eakin2010; @lugofernandez2010].

**Coral disease outbreaks (local to regional, contested):** Coral disease outbreaks may also act in synergy with other drivers by weakening coral resilience to other shocks, and the same works the other way around. For example, warmer water temperature and nutrient inputs have been correlated with increases in disease outbreaks such as white band syndrome [@bruno2007; @houk2010]. However, in cases where disease probability is density dependent, outbreaks are less likely to occur. This depends on the disease dynamics, transmission patterns, and contextual factors, which is why it is considered a contested driver.

**Low tides (local, well-established):** Low tides have been reported to induce mass coral mortality [@norstrom2009]. They expose the coral to higher solar radiation, luminosity, and temperature.

**Oil spills (local, controversial):** Oil spills can harm coral reefs by altering water chemistry and exposing corals to toxic hydrocarbons, which may cause tissue death, impaired feeding and calcification, changes in zooxanthellae productivity, and increased coral mortality. They can also reduce reef recovery by damaging reproduction, causing larval death, premature larval expulsion, and impaired larval settlement and recruitment [@fernandes2023].

**Shipwrecks and metal debris (local, well-established):** Shipwrecks and other large ferrous metal structures on coral reefs have been associated with localised phase shifts to corallimorph dominance. One hypothesis is that iron leaching from these structures enhances the growth of zooxanthellate corallimorpharians in iron-limited oligotrophic waters, enabling them to rapidly overgrow surrounding corals. This highlights rapid shipwreck removal as a management priority, particularly on remote reefs where background iron availability is low [@work2008; @davis2018].

##### Direct external drivers

**Fishing (local, well-established):** Overfishing reduces herbivory, leading to increasing macroalgal abundance due to reduced grazing [@mumby2007]. Fishing can also reduce the diversity of herbivores. When coral cover is reduced, there is increased colonization by algae, which in turn inhibits coral recruitment — i.e., positive feedback exists [@mumby2007; @norstrom2009]. When the abundance of fish is reduced, the ability to react to algal growth peaks is reduced as well, and mats of algae may establish and cover portions of a coral reef.

**Atmospheric CO₂ and ocean acidification (global, well-established):** Ocean acidification reduces the concentration of carbonate ions in the oceans, a fundamental compound required for corals to grow. With less carbonate accretion corals cannot fix their skeletons, hence the coral reef structure will be reduced. Many corals are expected to be unable to cope with CO₂ concentrations above 480 ppm and temperature increase above +2°C [@hoeghguldberg2007].

**Global warming (global, well-established):** Rising sea surface temperatures drive increased bleaching frequency and severity and interact with ocean acidification to compound physiological stress on corals.

##### Indirect external drivers

**Deforestation and land-use change (regional, proposed):** Terrestrial deforestation increases nutrient and sediment runoff to coastal waters. Elevated nutrients stimulate algal growth while suppressing coral growth rates, and sedimentation smothers coral recruits and reduces light availability [@hoeghguldberg2007].

**Coastal urbanization (local, proposed):** Coastal development increases sewage and runoff inputs, producing similar effects to deforestation through elevated nutrient and sediment loading.

**Rainfall and terrestrial runoff (regional, well-established):** Periods of heavy rainfall deliver pulses of sediment, nutrients, and freshwater to coastal reefs, which can suppress coral growth, depress octocoral recruitment, and promote coral disease outbreaks. Long-term monitoring has found annual rainfall to be one of the most consistent environmental correlates of community change on Caribbean reefs, influencing both scleractinian and octocoral assemblages [@tsounis2017].

**Food demand and fishing pressure (global, well-established):** Growing global food demand drives increased fishing effort, facilitated by market connectivity, technological development, and governance failures that allow overexploitation of reef-associated fish stocks [@deutsch2011].

##### Slow internal changes

**Herbivory decline (local, well-established):** Herbivorous fish regulate algal biomass through grazing, scraping, and bioerosion. Once macroalgae exceed a threshold size, they become unpalatable to many herbivores, establishing reinforcing feedback that favors continued algal dominance [@hughes2010]. Reduction in herbivore diversity and abundance diminishes the functional redundancy needed to buffer against episodic algal blooms following disturbance [@graham2024].

**Connectivity loss (regional, well-established):** Spatial connectivity enables larval exchange among reef populations, increasing genetic diversity and replenishment potential. Where connectivity is reduced, reefs rely on self-seeding and become more vulnerable to local population depletion, bleaching mortality, and recruitment failure [@hughes2010]. This is particularly relevant for isolated reef systems, where larval supply is largely locally derived.

##### Key thresholds

The primary thresholds are from temperature and carbon dioxide levels. However, because warming interacts with heatwaves, acidification, deoxygenation, sea-level rise, disease, pollution, and recovery lags, the effective tipping threshold may be lower than the central 1.2°C estimate [@pearcekelly2024].

- Ocean warming: 1.2°C global warming, range 0.7–1.5°C; also framed as 350 ppm CO₂, range 326–400 ppm
- Marine heatwaves / bleaching: 8–12 Degree Heating Weeks (DHW)
- Coral reef ecosystem collapse: live coral cover falling below roughly 10%, with prolonged failure to recover
- Ocean acidification: 538–572 ppm CO₂
- Sea level rise: around 8 mm/year, range 4–9 mm/year
- Deoxygenation: Coral homeostasis impaired around 2 mg/L O₂; lethal range 0.5–2 mg/L
- Storm intensity: Category 4 storms with a return time of ≤5 years / insufficient recovery time
- Pollution / eutrophication: Around 0.45 μg/L chlorophyll, marked with uncertainty in the table
- Disruption / land use: Agricultural/urban land use threshold is around 50% within 50 km of reefs
- Disease: Occurring during a time when temperatures are 1–2°C above ambient seasonal highs
##### Feedback mechanisms

The feedback maintaining each reef regime reflects competitive relationships among corals, macroalgae, turf algae, corallimorpharians, soft corals, sponges, and urchins. Each regime tends to reinforce itself by occupying space, modifying local conditions, and consuming resources in ways that disadvantage competing groups. However, the strength and permanence of these feedback mechanisms vary considerably, and long-term empirical data increasingly show that coral phase shifts can be bidirectional [@crisp2022; @graham2024].

##### Coral-dominated reefs

**Zooxanthellae symbiosis feedback (local, well-established):** The coral regime is reinforced by the mutualistic symbiosis between coral polyps and endosymbiotic dinoflagellates (zooxanthellae). Zooxanthellae supply corals with photosynthetically fixed carbon — meeting up to 90% of coral energy requirements — while receiving inorganic nutrients and a protected intracellular environment in return. This symbiosis underpins coral calcification rates, growth, and reproductive output. When coral cover is reduced by external disturbances, competitors can rapidly colonize available substrate, suppressing coral recruitment and weakening this reinforcing feedback.

**Structural complexity feedback (local, well-established):** High coral cover generates three-dimensional reef structure that supports diverse fish and invertebrate communities, which in turn provide herbivory, predation, bioerosion, and nutrient cycling that maintain conditions favorable to coral dominance. Pre-disturbance structural complexity has been identified as one of the strongest predictors of post-bleaching reef trajectory, with more complex reefs significantly less likely to shift to macroalgal dominance [@graham2015]. Structural degradation is therefore self-reinforcing, as reduced complexity diminishes the ecological processes that support coral recovery.

**Herbivore biomass feedback (local, well-established):** On recovering reefs, high fish biomass maintains low algal cover by clearing substrate for coral settlement and suppressing algal overgrowth of newly settled recruits [@pearcekelly2024].

**Thermal adaptation feedback (local to regional, emerging):** There is growing evidence that coral populations exposed to repeat heat stress events may develop increased thermal tolerance through selective mortality of heat-sensitive genotypes, phenotypic acclimation, and shifts in symbiont communities. Reefs that have survived prior bleaching events have in some cases, shown faster recovery and greater resistance during subsequent events, suggesting this feedback may strengthen over successive disturbances [@graham2024].

##### Algae-dominated reefs

**Unpalatability feedback (local, well-established):** Once macroalgae reach a sufficient size and biomass, they become physically and chemically unpalatable to many herbivorous fish, reducing grazing pressure and enabling further algal growth and recruitment [@hughes2010]. Allelopathic interactions, whereby macroalgae release chemical compounds that inhibit coral larval settlement and growth, further compound this effect, suppressing coral recovery on regime-shifted reefs.

**Coral recruitment inhibition feedback (local, well-established):** Macroalgal cover physically excludes coral larvae from suitable settlement substrate and, through shading and allelopathy, reduces post-settlement coral survival.

**Nutrient cycling feedback (local, proposed):** Macroalgal-dominated reefs retain nutrients within algal biomass in ways that further favor algal growth over coral growth. Lower carbon-to-nitrogen ratios in macroalgal tissue, indicative of higher ambient nutrient loads, have been associated with a significantly greater probability of regime shifts [@burkepile2013; @graham2015].

**Turf algae feedback (local, well-established):** Dense turf mats trap sediment, reduce substrate availability for coral larvae, and can transition to macroalgal dominance under elevated nutrient conditions. Coral-to-turf transitions are the second most documented form of phase shift globally, accounting for approximately 14% of reported cases [@crisp2022].

##### Corallimorpharians, soft corals, and sponges

**Competitive exclusion feedback (local, well-established):** Corallimorpharians, soft corals, and sponges compete with hard corals for space and light. Following strong disturbance events that reduce hard coral populations, these functional groups can rapidly colonize available substrate, suppressing hard coral recruitment through overgrowth, shading, and chemical interference, locking the system into a non-calcifying alternative state [@norstrom2009; @crisp2022].

##### Urchin barrens

**Bioerosion feedback (local, well-established):** Urchin barrens typically develop in environments with low predator diversity and abundance. At high densities, urchins shift from grazing algae to actively bioeroding the reef framework, including live coral tissue. The urchin barren regime is self-reinforcing when bioerosion rates exceed reef accretion rates, progressively reducing structural complexity and further suppressing the coral diversity that would otherwise support predator recovery [@norstrom2009].

#### Impacts on ecosystem services and human well-being

Coral reefs occupy less than 0.5% of the ocean floor yet support approximately 25% of all marine species and provide a disproportionate share of the ecosystem services upon which hundreds of millions of people depend, particularly in tropical coastal regions [@moberg1999; @hoeghguldberg2019]. Regime shifts away from coral-dominated states result in the degradation or loss of multiple interacting services simultaneously, with cascading consequences for human well-being that extend well beyond the reef itself.

##### Provisioning services

**Fisheries and food security:** Coral reefs directly support the food security and livelihoods of an estimated 500 million people globally, with reef-associated fisheries providing a critical source of protein in many small island developing states and coastal communities across the Indo-Pacific, Caribbean, and Indian Ocean [@hoeghguldberg2019]. Regime shifts to macroalgal, soft coral, or sponge dominance reduce fish biomass and alter community composition, with the greatest losses typically occurring among coral-dependent species (corallivores and small-bodied planktivores) that are among the first to decline following coral loss [@graham2015]. The short-term fisheries impact of bleaching may be partially buffered by compensatory increases in herbivore productivity, though such gains are likely unsustainable if coral cover does not recover and if herbivorous fish continue to be heavily targeted by fisheries.

**Wild animal and plant products:** Beyond finfish, coral reefs support diverse invertebrate fisheries (including sea cucumbers, lobsters, mollusks, and sea urchins), many of which are disproportionately important to subsistence and small-scale commercial fishing in reef-adjacent communities. These species assemblages are sensitive to regime shifts, particularly those involving structural collapse or urchin barren formation.

##### Regulating services

**Coastal protection:** Structurally complex coral reefs play a critical role in everyday shoreline maintenance by dissipating routine wave energy before it reaches shore, limiting chronic erosion. This function depends strongly on the reef’s three-dimensional structure. When reefs undergo regime shifts (especially after severe bleaching) their carbonate budgets can shift to net erosion, meaning the reef framework begins to break down rather than grow or be maintained [@graham2015; @januchowskihartley2017]. As structural complexity declines, the reef’s ability to buffer wave energy is reduced, increasing day-to-day coastal exposure and accelerating shoreline retreat.

**Water purification and biogeochemical cycling:** Coral reef ecosystems contribute to water quality through the cycling of nutrients, particulate matter, and dissolved organic compounds. Sponge-dominated alternative states alter the nature and efficiency of these processes. Shifts to algal dominance can promote hypoxic conditions and alter microbial community composition in ways that further degrade water quality, creating feedbacks that disadvantage coral recovery.

**Natural hazard regulation:** Beyond routine wave buffering, coral reefs also reduce the exposure of coastal communities to extreme episodic events such as tsunamis, tropical cyclones, and associated storm surges. The value of this protection has been estimated in the tens of billions of dollars annually at the global scale, with particularly severe consequences for low-lying islands and coastal communities that have limited capacity to adapt. Its loss through reef degradation represents a major increase in coastal risk that is rarely fully captured in economic assessments of reef value [@spalding2017].

##### Cultural services

**Recreation and tourism:** Coral reef tourism generates an estimated US$36 billion annually and supports livelihoods across more than 100 countries [@spalding2017]. The Great Barrier Reef alone contributes approximately US$6.4 billion per year and supports over 64,000 jobs, predominantly through tourism [@deloitte2017]. Regime shifts substantially reduce the experiential value of reef tourism: macroalgal or sponge-dominated reefs attract fewer visitors and command lower willingness-to-pay among divers and snorkelers.

**Aesthetic and spiritual values:** Coral reefs hold profound aesthetic, cultural, and spiritual significance for many reef-adjacent communities, particularly indigenous and traditional communities in the Pacific, Indian Ocean, and Caribbean. Traditional ecological knowledge systems, customary resource management practices, and cultural identities are often deeply intertwined with reef health. The loss of these values through regime shifts represents a form of cultural harm that is poorly captured by economic valuation frameworks [@moberg1999; @hoeghguldberg2019].

**Knowledge and educational values:** Coral reefs represent irreplaceable systems for ecological research. The long-term monitoring datasets that have generated the most important insights into reef resilience and regime shift dynamics themselves represent significant scientific assets whose value depends on the continued existence of the reef systems they monitor.

##### Human well-being

**Food and nutrition:** The loss of reef fisheries disproportionately affects communities with limited dietary alternatives, increasing vulnerability to malnutrition and food insecurity. In small island developing states, per capita fish consumption from reef fisheries can exceed 50 kg per year, and declines in reef fish biomass following regime shifts translate directly into reduced dietary protein availability [@hoeghguldberg2019].

**Livelihoods and economic activity:** Reef-dependent livelihoods, including fishing, tourism, aquaculture, and coastal construction, are directly threatened by regime shifts. The impacts are highly unequally distributed, falling most heavily on communities with the least capacity to diversify or adapt.

**Security of housing and infrastructure:** Coastal protection provided by structurally complex reefs reduces the cost and frequency of storm damage to housing and infrastructure in reef-adjacent communities. As reefs degrade and lose their wave-attenuation capacity, these communities face escalating infrastructure risks that compound other dimensions of climate vulnerability.

##### Economic valuation

Early global estimates valued coral reef ecosystem services at approximately US$6,075 per hectare per year [@costanza1997]. More recent assessments suggest substantially higher values, averaging on the order of US$80,000–350,000 per hectare per year, and exceeding US$1 million per hectare per year in high-value contexts such as tourism and coastal protection [@costanza2014; @degroot2012; @spalding2017]. These aggregate figures, however, mask enormous spatial variation: values differ by several orders of magnitude across locations, with the highest concentrations where reefs are both ecologically intact and proximate to dense human populations or high-value tourism markets [@fezzi2022]. Many of these services are substantially reduced or lost entirely when regime shifts occur, though the economic costs of such shifts are rarely fully quantified in management decision-making frameworks.

#### Management options

##### Options for enhancing resilience

“The persistence of hard coral-dominated reefscapes beyond 2050 will be heavily reliant on 2 things: the ability of corals to increase their upper thermal bleaching limits by ~0.1°C per decade, and management that produces local conditions that constrain excessive algal biomass proliferation during inter-disturbance intervals.” [@wooldridge2005]

Coral reef regime shifts are driven by local, regional, and global drivers. Given that temperature and acidification are global drivers that are difficult for local managers to influence, Hoegh-Guldberg et al. [-@hoeghguldberg2007] emphasize the management of local-level drivers. Due to the multi-causal nature of coral regime shifts, scholars emphasize the necessity of managing coral reefs using an ecosystem approach. Such an approach requires considering the interaction between land and sea, as well as the scale and origin of the stressors when making decisions at different scales of governance [@moberg1999]. Management programs that successfully address the improvement of water quality, reduction of sediments, nutrients, toxins, pathogens, and fishing pressure will increase the likelihood of corals recovering from shocks like bleaching events [@wooldridge2005; @hoeghguldberg2007; @houk2010]. The rapid removal of shipwrecks and other large metal debris from reefs has also been identified as an important management priority, given the potential for such objects to trigger localized phase shifts to corallimorph-dominated states [@work2008].

Herbivore populations are a key driver that can be actively managed because reduced grazing can increase the vulnerability to regime shifts [@mumby2007]. For example, herbivorous fish can be protected. It has been suggested that markets should be transformed to incorporate a body of incentives to prevent the depletion of species in critical functional groups [@bellwood2004]. The abundance of sea urchins should also be carefully managed because urchin dominance can produce negative effects on coral recruitment [@norstrom2009].

The spatial resilience of coral reefs is a regional driver that can be managed by managing connectivity, metapopulation dynamics, and by taking into consideration the spatial distribution of coral reefs [@moberg1999; @nystrom2001]. This is because large-scale regional shifts are typically preceded by smaller-scale localized shifts. Therefore, monitoring the occurrence and spatial distribution of smaller-scale regime shifts may help to anticipate, and potentially avert, large-scale catastrophic shifts [@norstrom2009].

To manage spatial resilience, Bellwood et al. [-@bellwood2004] recommend increasing the rate of establishment and size of no-take areas, including “cool-spots” of biodiversity. The reason is that areas with low species richness may be more vulnerable, as they may have lost functional groups or may have low functional redundancy. Hence, minor changes in such ecosystems may trigger regime shifts locally and erode spatial resilience regionally. However, no-take marine reserve status alone has not been shown to reliably prevent climate-driven regime shifts, indicating that protection from fishing must be combined with broader management of water quality, herbivory, and habitat structure to be effective [@graham2015]. International agreements are also badly needed to keep ocean conditions below the 480 ppm and +2°C thresholds for coral reefs [@hoeghguldberg2007].

Reducing pressure from global-scale drivers requires coordinated global action but would substantially increase the resilience of coral reefs.

#### Key references
